---
title: "Introduction to R"
author: "Gamaliel Cabria"
date: "`r Sys.Date()`"
output: 
  github_document:
    keep_md: true
---

# A Quick Introduction to *R*
R is a powerful programming language and environment for statistical computing and graphics. It is widely used among statisticians and data miners for data analysis and visualization. 

In base R (without additional packages), you can create variables, manipulate data, and perform calculations with simple commands. Here are some basic examples:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Create a variable
x <- 5
# Perform a calculation
y <- x * 2
# Print the result
print(y)  
# Create a vector
v <- c(1, 2, 3, 4, 5)
# Calculate the mean of the vector
mean_v <- mean(v)
# Print the mean
print(mean_v)  
```

Think of R as a more complex and **powerful calculator** that can handle complex data analysis tasks. It allows you to perform operations on data, create visualizations, and apply statistical models.

# R as programming language
## R Data Structures
R has several built-in data structures that are essential for data analysis. Data structures in R are ways to organize and store data so that it can be easily manipulated and analyzed. Each structure has its own characteristics and is suited for different types of data.

The most common ones include:

### Core Data Structures in Base R

| **Structure**   | **Description**                                                                 | **Example**                                                                 |
|-----------------|----------------------------------------------------------------------------------|------------------------------------------------------------------------------|
| **Vector**      | One-dimensional sequence of elements of the same type                           | `x <- c(1, 2, 3, 4)`<br>`names <- c("Alice", "Bob")`                         |
| **Matrix**      | Two-dimensional array of the same type                                          | `m <- matrix(1:6, nrow = 2)`<br>`m[1, 2]  # Access element`                  |
| **Array**       | Multi-dimensional generalization of a matrix                                    | `a <- array(1:12, dim = c(2, 3, 2))`<br>`a[1, 2, 1]`                         |
| **Data Frame**  | Table with columns of different types                                           | `df <- data.frame(name=c("A", "B"), age=c(25, 30))`<br>`df$age`             |
| **List**        | Collection of elements of any type                                              | `lst <- list(name="Alice", age=25, scores=c(90, 85))`<br>`lst[[2]]`         |
| **Factor**      | Categorical variable with predefined levels                                     | `f <- factor(c("low", "medium", "high"))`<br>`levels(f)`                    |
| **Environment** | Collection of objects (variables, functions) in a specific scope              | `env <- new.env()`<br>`assign("x", 10, envir = env)`<br>`env$x`              |
| **Function**    | A block of code that can be called with arguments                               | `my_func <- function(x) { x^2 }`<br>`my_func(3)`                              |

## R Data Types and Variables
Like any programming language, R allows you to create variables to store data. Variables can hold different types of data, such as numbers, text, or logical values. Here are some examples of creating and using variables in R:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Numeric type
num_var <- 3.14
print(num_var)        

# Integer type
int_var <- 5L  # 'L' suffix indicates an integer
print(int_var)        

# Character type
char_var <- "Hello, R!"
print(char_var)      

# Logical type
log_var <- TRUE
print(log_var)        

# Complex type
complex_var <- 1 + 2i  # 'i' indicates an imaginary part
print(complex_var)    

# Raw type
raw_var <- charToRaw("A")  # Converts character to raw bytes
print(raw_var)     

# Factor type
factor_var <- factor(c("low", "medium", "high"))  # Categorical data
print(factor_var)
print(levels(factor_var)) 

# Date type
date_var <- as.Date("2025-01-01")  # Date object
print(date_var)      

# POSIXct type
posix_var <- as.POSIXct("2025-01-01 12:00:00")  # Date-time object
print(posix_var)

# Function type
func_var <- function(a, b) a + b  # User-defined function
print(func_var(2, 3))  

# List type
list_var <- list(name="Gama", age=18, scores=c(96, 88))  # Heterogeneous container
print(list_var)  

# Data frame type
df_var <- data.frame(a=1:4, b=c("A", "B","C","Z"))  # Table-like structure
print(df_var)  

# Matrix type
matrix_var <- matrix(1:4, nrow=2)  # 2D numeric structure
print(matrix_var)  

# Array type
array_var <- array(1:8, dim=c(2, 2, 2))  # Multi-dimensional structure
print(array_var)  
```

R is also an **interactive language**, meaning you can type commands directly into the console and see results immediately. This makes it easy to experiment with data and perform quick calculations.

### Summary of R Data Types and Structures
| **Type**       | **Description**                                 | **Example**                                   | **typeof()**     | **class()**     |
|----------------|-------------------------------------------------|-----------------------------------------------|------------------|-----------------|
| **Numeric**     | Numbers with decimal points                     | `x <- 3.14`                                    | `"double"`       | `"numeric"`     |
| **Integer**     | Whole numbers (explicit with `L`)               | `x <- 5L`                                      | `"integer"`      | `"integer"`     |
| **Character**   | Text strings                                    | `x <- "hello"`                                 | `"character"`    | `"character"`   |
| **Logical**     | Boolean values: TRUE or FALSE                   | `x <- TRUE`                                    | `"logical"`      | `"logical"`     |
| **Complex**     | Complex numbers                                 | `x <- 1 + 2i`                                  | `"complex"`      | `"complex"`     |
| **Raw**         | Raw bytes (rarely used)                         | `x <- charToRaw("A")`                          | `"raw"`          | `"raw"`         |
| **Factor**      | Categorical data with levels                    | `x <- factor(c("low", "high"))`               | `"integer"`      | `"factor"`      |
| **Date**        | Date objects                                    | `x <- as.Date("2025-01-01")`                   | `"double"`       | `"Date"`        |
| **POSIXct**     | Date-time object                                | `x <- as.POSIXct("2025-01-01 12:00:00")`       | `"double"`       | `"POSIXct"`     |
| **Function**    | User-defined or built-in functions              | `x <- function(a, b) a + b`                    | `"closure"`      | `"function"`    |
| **List**        | Heterogeneous container                         | `x <- list(name="Gama", age=30)`              | `"list"`         | `"list"`        |
| **Data Frame**  | Table-like structure                            | `x <- data.frame(a=1:2, b=c("A", "B"))`        | `"list"`         | `"data.frame"`  |
| **Matrix**      | 2D numeric structure                            | `x <- matrix(1:4, 2, 2)`                       | `"integer"`      | `"matrix"`      |
| **Array**       | Multi-dimensional structure                     | `x <- array(1:8, c(2,2,2))`                    | `"integer"`      | `"array"`       |
| **Environment** | Collection of objects in a scope                | `x <- new.env()`                               | `"environment"`  | `"environment"` |

{:.note}
>If you do not know data type or structure of an object, you can use the `typeof()` and `class()` functions to check it. For example:
```{r, echo=TRUE, warning=FALSE, message=FALSE} 
typeof(x)  # Check the type of x
class(x)   # Check the class of x
```

# R Basics

## Basic Syntax
R syntax is straightforward and similar to other programming languages. Here are some basic elements of R syntax:
- **Comments**: Use `#` to add comments in your code.
- **Assignment**: Use `<-` or `=` to assign values to variables.
- **Functions**: Call functions using their names followed by parentheses, e.g., `mean(x)`.
- **Operators**: Use operators like `+`, `-`, `*`, `/` for arithmetic operations, and `==`, `!=`, `<`, `>` for comparisons.
- **Control Structures**: Use `if`, `else`, `for`, and `while` for control flow.
- **Packages**: Use `library(package_name)` to load additional functionality.

## Indexing and Subsetting
In R, you can access elements of vectors, matrices, data frames, and lists using indexing. Here are some examples:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Accessing elements in a vector
v <- c(10, 20, 30, 40)
print(v)
print(v[2])  

# Accessing elements in a matrix
m <- matrix(1:9, nrow=3)
print(m)
print(m[2, 3])  

# Accessing elements in a data frame
df <- data.frame(name=c("Alice", "Bob"), age=c(25, 30))
print(df)
print(df[1, ])  
print(df[1, "name"])  
print(df$name)  

# Accessing elements in a list
lst <- list(a=1:5, b="Hello", c=TRUE)
print(lst)
print(lst$b)  

# Accessing elements in a factor
f <- factor(c("low", "medium", "high"))
print(f[2])  

# Accessing elements in a character vector
char_vec <- c("apple", "banana", "cherry")
print(char_vec[1])  
```

{:.note}
>Indexing in R is **1-based**, meaning the first element is accessed with index 1. You can also use **negative indices** to exclude elements or logical vectors to filter data.

## Vectorized Operations
R is designed for vectorized operations, meaning you can perform operations on entire vectors or matrices without using explicit loops. This makes R efficient for data manipulation. Here are some examples:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Vectorized addition
v1 <- c(1, 2, 3)
v2 <- c(4, 5, 6)
result <- v1 + v2  # Adds corresponding elements
print(result)  
# Vectorized multiplication
v3 <- c(2, 3, 4)
result2 <- v1 * v3  # Multiplies corresponding elements
print(result2)  
# Vectorized comparison
v4 <- c(1, 2, 3, 4)
result3 <- v4 > 2  # Compares each element to 2
print(result3)  
# Matrix operations
m1 <- matrix(1:4, nrow=2)
m2 <- matrix(5:8, nrow=2)
result4 <- m1 + m2  # Adds corresponding elements
print(m1)
print(m2)
print(result4)  
# Vectorized functions
v5 <- c(1, 2, 3, 4)
result5 <- sqrt(v5)  # Applies square root to each element
print(result5)  
# Vectorized logical operations
v6 <- c(TRUE, FALSE, TRUE)
v7 <- c(FALSE, TRUE, TRUE)
result6 <- v6 & v7  # Logical AND operation
print(result6)  
```

Vectorized operations allow you to write concise and efficient code, **avoiding the need for explicit loops** like in other languages. This is one of the key features that makes R powerful for data analysis.

## Recycling Rules *R*
In R, when performing operations on vectors of different lengths, R applies a concept called **recycling**. This means that if one vector is shorter than the other, R will repeat the shorter vector until it matches the length of the longer vector. Here are some examples:
```{r, echo=TRUE, warning=TRUE, message=FALSE}
# Example with different lengths
v1 <- c(1, 2, 3)
v2 <- c(10, 20)
result <- v1 + v2  # The second vector is recycled
print(result)  

# Example with recycling in a matrix
m1 <- matrix(1:6, nrow=2)
m2 <- c(10, 20)  # Shorter vector
result2 <- m1 + m2  # The second vector is recycled
print(m1)
print(m2)
print(result2)  
```

## Logical Operations in **R**
Logical operations in R are used to perform comparisons and logical tests on data. They return boolean values (`TRUE` or `FALSE`) based on the conditions specified. Here are some common logical operations:

{:.note}
>'TRUE' and 'FALSE' are written in uppercase in R, and they are used to represent boolean values. 'TRUE' and 'FALSE' can be represented as 1 and 0, respectively, but are not strict equivalents.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Logical operations
# Create two vectors
v1 <- c(1, 2, 3, 4)
v2 <- c(3, 4, 3, 6)
# Comparison operators
equal <- v1 == v2  # Check if elements are equal
print(equal)  
greater_than <- v1 > v2  # Check if elements in v1 are greater than those in v2
print(greater_than)  
less_than <- v1 < v2  # Check if elements in v1 are less than those in v2
print(less_than)  
# Logical operators
and <- v1 > 2 & v2 < 5  # Logical AND operation
print(and)  
or <- v1 < 2 | v2 > 5  # Logical OR operation
print(or)  
not <- !(v1 < 3)  # Logical NOT operation
print(not)  
# Combining logical conditions
combined_condition <- (v1 > 2) & (v2 < 5)  # Combine conditions with AND
print(combined_condition)  
# Logical indexing
# Use logical vectors to index elements
v3 <- c(10, 20, 30, 40)
logical_index <- v3 > 20  # Create a logical vector
print(logical_index)  
# Use logical vector to subset the original vector
subset_v3 <- v3[logical_index]  # Subset using logical indexing
print(subset_v3)  
```

### Conditional Statements
Conditional statements in R allow you to execute code based on certain conditions. The most common conditional statements are `if`, `else if`, and `else`. Here are some examples:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Conditional statements
x <- 5

# If-else statement
if (x > 0) {
  print("x is positive")
} else if (x < 0) {
  print("x is negative")
} else {
  print("x is zero")
}

# Nested if-else statement
if (x > 0) {
  print("x is positive")
} else {
  if (x < 0) {
    print("x is negative")
  } else {
    print("x is zero")
  }
}

# Switch statement
fruit <- "apple"
switch(fruit,
       apple = "This is an apple",
       banana = "This is a banana",
       orange = "This is an orange",
       "Unknown fruit")  # Default case if no match
```

Conditional statements are essential for controlling the flow of your R programs, allowing you to execute different code blocks based on specific conditions.

## Understanding Missing Values or **NA** in R
In R, missing values are represented by `NA` (Not Available). They indicate that a value is not present or is unknown. Handling missing values is crucial in data analysis, as they can affect calculations and results. 

There are several ways of working with missing values in R:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Create a vector with missing values
v <- c(1, 2, NA, 4, NA, 6)
mean(v)
mean(v, na.rm = TRUE)  # Calculate mean while ignoring NA values

# Check for missing values
is.na(v)  # Returns TRUE for NA values

# Count missing values
na_count <- sum(is.na(v))  # Count the number of NA values
print(na_count)

# Remove missing values
v_no_na <- na.omit(v)  # Remove NA values from the vector
print(v_no_na)

# Replace missing values
v[is.na(v)] <- 0  # Replace NA values with 0
print(v)  # Print the vector after replacing NA values
```

Missing values can occur in various data types, including vectors, matrices, data frames, and lists. It is essential to handle them appropriately to ensure accurate analysis and results.

## Reading and Writing Data
There are several ways to read and write data in R, depending on the format of the data. 
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Let us create a sample data frame
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Score = c(90, 85, 95)
)
# Writing data to a CSV file
write.csv(df, "./sample_data.csv", row.names = FALSE)  # Save data frame to CSV file
# Reading data from a CSV file
df_read <- read.csv("./sample_data.csv")  # Read data from CSV file
print(df_read)  # Print the read data frame
```

# 🧠 Extra Base R information
## The HELP System
R has a built-in help system that provides documentation for functions, packages, and datasets. You can access help in several ways:
```{r}
?mean
help("lm")
example(mean)

# Try this too:
help(package = "stats")  # Get help for the 'stats' package
help.search("linear model")  # Search for help on linear models
```

## Apply family functions
The apply family functions in base R are a set of functions used to perform loop-like operations over elements of data structures (vectors, lists, matrices, etc.) in a more concise and often more efficient way than using explicit for loops.
Examples of apply family functions include `apply()`, `lapply()`, `sapply()`, `tapply()`, and `mapply()`. These functions allow you to apply a function to each element or subset of a data structure without writing explicit loops.
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Using apply family functions
# Create a matrix
m <- matrix(1:9, nrow=3)
# Apply function to rows
row_sums <- apply(m, 1, sum)  # Sum of each row
print(row_sums)  
# lapply
lst <- list(a=1:5, b=6:10, c=11:15)
lapply_result <- lapply(lst, mean)  # Calculate mean of each list element
print(lapply_result)  
#sapply
sapply_result <- sapply(lst, mean)  # Simplified version of lapply
print(sapply_result)  
# tapply
ages <- c(25, 30, 35, 40)
group <- c("A", "A", "B", "B")
tapply(ages, group, mean)  # Mean age by group
```


# Summary: Base R

In this introduction to **base R**, you've learned how to create and manipulate variables, explore core data types and structures, perform vectorized operations, handle missing values, and apply functions efficiently using the `apply` family. You've also seen how to read and write data, use conditional logic, and leverage R's built-in help system.

Base R provides a strong foundation for data analysis, and understanding its syntax and logic will help you confidently explore more advanced tools like the tidyverse, data visualization, and statistical modeling. With practice, R becomes a powerful ally in data-driven problem-solving.

Happy analysis!
